A parametricity-based formalization of semi-simplicial and semi-cubical sets

### 2023-03-19 paper/paper.tex
Preparing a response, answering several points, but not all yet

\section{Introduction}
In the context of functional programming, Reynolds' parametricity~\cite{reynolds83} interprets types as relations characterizing the observational behavior of programs of this type. Parametricity can be iterated, and it has been noted that iterated Reynolds' parametricity can be interpreted in various cubical sets~\cite{johann17,altenkirch15,moulin16,moeneclaey21,moeneclaey22phd}. We obtain a \emph{unary} variant of Reynolds' \emph{binary} parametricity by using predicates or families instead of relations or graphs, in which case, we obtain a form of realizability~\cite{bernardy12,lasson12,moulin16}. It has then been noted that iterated unary parametricity has an augmented simplicial flavor~\footnote{Private communication with Hugo Moeneclaey and Thorsten Altenkirch}.

@@

The outline of the paper is as follows. We recall the definition of the augmented semi-simplicial and semi-cubical categories using a combinatorial presentation, that directly leads to our generalization to $\nu$-sets, in \ref{sec:nu}. We then proceed to explain the connection between the indexed representation and parametricity in \ref{sec:indexed-param}. Before spelling out the details of our formalization in type-theoretic language in \ref{sec:itt}, we build up intuitions in \ref{sec:intuition}. We end with some of the finer details of our mechanization in \ref{sec:mech}.

@@

The standard augmented $0$-semi-simplex is a singleton of one color. Standard augmented $n$-semi-simplices have a geometric interpretation, and we illustrate them for dimensions $1$, $2$, and $3$.

\begin{example}[$\DeltaPlus^1$]
  In dimension $1$, the standard augmented semi-simplex can be pictured as a point, colored black, corresponding to the unique morphism in $\Hom(0, 1)$. This point is the identity in $\Hom(1, 1)$; it is hence shown as a singleton $\kstar$.
  \begin{equation*}
    \begin{tikzcd}
      \kstar
    \end{tikzcd}
  \end{equation*}
\end{example}

@@

\subsection{Semi-cubical sets}
Semi-cubical sets are defined like augmented semi-simplicial sets except that $\DeltaPlus$ is replaced by $\Cube$ in which we take sequences of $L$, $R$ and $\star$, instead of sequences of $0$ and $\kstar$.

@@

Let us apply this correspondence to the case of semi-cubical sets. A semi-cubical set can be represented as a sequence of fibrations, together with appropriate coherence conditions.
\begin{equation*}
  \begin{tikzcd}
    X_0: \U & X_1: \U \arrow[l, "\delta^L" description, shift left=2] \arrow[l, "\delta^R" description, shift right=2] & X_2: \U \arrow[l, "\delta^{L\kstar}" description, shift left=6] \arrow[l, "\delta^{R\kstar}" description, shift left=2] \arrow[l, "\delta^{\kstar R}" description, shift right=2] \arrow[l, "\delta^{\kstar L}" description, shift right=6] & \ldots
  \end{tikzcd}
\end{equation*}

If we think of $X_1$ fibered over two copies of $X_0$ in an indexed way, we get $X_1$ of type $X_0 \rightarrow X_0 \rightarrow \U$. Continuing the process, we can think of $X_2$ fibered over four copies of $X_1$ themselves fibered over two copies of $X_0$ as a dependent type. We have a priori a dependency over eight copies of $X_0$ but these are actually four due to the coherence conditions. Based on these coherence conditions, the types of $X_0, X_1, X_2$ happen to have the form:
\begin{align*}
  X_0 & :              & \U                                                     \\
  X_1 & :              & X_0 \rightarrow X_0 \rightarrow \U                     \\
  X_2 & : \Pi a b c d. & X_1(a)(b) \rightarrow X_1(a)(c) \rightarrow X_1 (b)(d) \\
      &                & \rightarrow X_1 (c)(d) \rightarrow \U                  \\
  \ldots
\end{align*}

@@

But first, we start with a section gradually and informally explaining the intuition of the construction.

@@

We assign every $X_n$ the type $\fullframe[n] \rightarrow \U$ uniformly, applying to the description above the isomorphism between $A \rightarrow B \rightarrow C$ and $A \times B \rightarrow C$, or between $\Pi a: A.\, (B a \rightarrow C)$ and $(\Sigma a: A.\, B a) \rightarrow C$. Let us now illustrate how we build $\fullframe[n]$. To begin, let us set $\fullframe[0] = \unittype$, so that the type $\U$ of $X_0$ can be equivalently formulated as $\unittype \rightarrow \U$. Then, more generally, we let each $\fullframe[n]$ consist of $n$ layers, written $\layer[n][p]$ with $p < n$, that we stack in order, starting from the $\unittype$ type and writing $\framep[n][p]$ for the $p$ first layers of a $\fullframe[n]$, so that $\fullframe[n]$ is $\framep[n][n]$. For instance, $X_1$ is made of one layer, so that it can be written as a $\Sigma$-type of an inhabitant of the $\unittype$ and $\layer[1][0]$, as shown in Fig.~\ref{fig:intuition}. Fig.~\ref{fig:intuition} also mentions how the type of $X_2$ is structured.

@@

So far, we have not paid attention to the fact that we have a dependent type, shown as a $\Sigma$. Let us be more precise about this requirement. First, $\fullframe[n]$ depends on all $X_i$ up to $n - 1$. So, we need to package up $X_i$, for $i < n$, into a type, which we call $\X[][<n][]$. This allows us to give the type $\X[][<n][] \rightarrow \U$ to $\fullframe[n]$. Then, for $D: \X[][<n][]$, representing an initial prefix of $X_0, X_1, \ldots X_{n - 1}$, the indexed set $X_n$ has type $\fullframe[n](D) \rightarrow \U$. Thus, $\framep[n][p]$, $\layer[n][p]$ and $\painting[n][p]$ also depend on $D$. We can then refine the previous equation by showing the dependencies on $D$. In particular, $X_n$ is just $D.\tl$.
\begin{align*}
   & \framep[n][p](D) \defeq                                                                                      \\
   & \;\;\;\;\Sigma a_n: (\ldots (\Sigma \unitpoint: \unittype.\, \layer[n][0](D)) \ldots) .\,\layer[n][p - 1](D) \\
   & \painting[n][p](D) \defeq                                                                                    \\
   & \;\;\;\;\Sigma l_p: \layer[n][p](D).\, (\ldots (\Sigma l_n: \layer[n][n - 1](D) .\, D.\tl) \ldots)
\end{align*}

@@

The definition is dispatched over tables \ref{tab:coind}, \ref{tab:core}, \ref{tab:frames}, \ref{tab:faces} and~\ref{tab:coh}. Table \ref{tab:coind} describes a $\nu$-set in indexed form, as a coinductive limit of $n$-truncated $\nu$-sets. The $n$-truncated $\nu$-sets are themselves described in \ref{tab:core}. The $n$th component is a family over a $\fullframe$. The type $\fullframe$ is recursively defined in \ref{tab:frames}, using the auxiliary definitions of $\layer$ and $\painting$. The type $\layer$ and $\painting$ are dependent over $\restrf$, and these restrictions are defined on \ref{tab:faces}. These restrictions are defined using auxiliary definitions of $\restrl$ and $\restrc$.

Notably, the definition of $\restrl$ relies on a definitional equality expressing the commutation of the composition of $\restrf$. This commutation is not provable by computation so we have to prove it propositionally before using the reflection rule. Proving this itself requires an induction on the dimension, and on the structure of $\framep$, $\layer$, and $\painting$. This is what $\cohf$ proves, as shown in the table \ref{tab:coh}, using auxiliary definitions $\cohl$ and $\cohc$. Even though it looks independent of the other tables, $\cohf$ has to be proved mutually with the definitions of $\framep$, $\layer$, $\painting$, and their corresponding restrictions. More precisely, for a fixed $n$, the block of $\framep$, $\restrf$, and $\cohf$ has to be mutually defined by induction on $p$. Also, each of $\painting$, $\restrc$, and $\cohc$ is built by induction from $p$ to $n$. The $\painting$ block at $n$ relies on the $\framep$ block at $n$, but, the converse dependency is only on lower $n$, so this is well-founded. Note that $\layer$, $\restrl$ and $\cohl$ are just abbreviations. We leave however implicit the exact way this mutual recursion can be formalized as this stage.

>>>>

\section{Introduction}
In the context of functional programming, Reynolds' parametricity~\cite{reynolds83} interprets types as relations characterizing the observational behavior of programs of this type. Parametricity can be iterated, and it has been noted that iterated Reynolds' parametricity has the structure of a cubical set~\cite{johann17,altenkirch15,moulin16,moeneclaey21,moeneclaey22phd}. We obtain a \emph{unary} variant of Reynolds' \emph{binary} parametricity by using predicates or families instead of relations or graphs, in which case, we obtain a form of realizability~\cite{bernardy12,lasson12,moulin16}. It has then been noted that iterated unary parametricity has the structure of an augmented simplicial set~\footnote{Private communication with Hugo Moeneclaey and Thorsten Altenkirch}.

@@

The outline of the paper is as follows. We recall the definition of the augmented semi-simplicial and semi-cubical categories using a combinatorial presentation, that directly leads to our generalization to $\nu$-sets, in \ref{sec:nu}. We then proceed to explain the connection between the indexed representation and parametricity in \ref{sec:indexed-param}. Before spelling out the details of our formalization in type-theoretic language in \ref{sec:ett} and \ref{sec:itt}, we build up intuitions in \ref{sec:intuition}. We end with some of the finer details of our mechanization in \ref{sec:mech}.

@@

The standard augmented $0$-semi-simplex is a singleton made of one color (in our case black). Standard augmented $n$-semi-simplices for $n\geq 1$ have a geometric interpretation, and we illustrate them for dimensions $1$, $2$, and $3$.

\begin{example}[$\DeltaPlus^1$]
  The standard augmented $1$-semi-simplex can be pictured as a point, colored black, corresponding to the unique morphism in $\Hom(0, 1)$. This point is the identity in $\Hom(1, 1)$; it is hence shown as a singleton $\kstar$.
  \begin{equation*}
    \begin{tikzcd}
      \kstar
    \end{tikzcd}
  \end{equation*}
\end{example}

@@

\subsection{Semi-cubical sets}
Semi-cubical sets are defined like augmented semi-simplicial sets except that $\DeltaPlus$ is replaced by $\Cube$ in which we take sequences of $L$, $R$ and $\star$, instead of sequences of $0$ and $\kstar$. They represent ordinary semi-cubical sets, with faces (and no augmentation).

@@

Let us apply this correspondence to the case of semi-cubical sets. A semi-cubical set can be represented as a sequence of fibrations, together with appropriate coherence conditions.
\begin{equation*}
  \begin{tikzcd}
    X_0: \U & X_1: \U \arrow[l, "\delta^L" description, shift left=2] \arrow[l, "\delta^R" description, shift right=2] & X_2: \U \arrow[l, "\delta^{L\kstar}" description, shift left=6] \arrow[l, "\delta^{R\kstar}" description, shift left=2] \arrow[l, "\delta^{\kstar L}" description, shift right=2] \arrow[l, "\delta^{\kstar R}" description, shift right=6] & \ldots
  \end{tikzcd}
\end{equation*}

If we think of $X_1$ fibered over two copies of $X_0$ in an indexed way, we get $X_1$ of type $X_0 \rightarrow X_0 \rightarrow \U$. Continuing the process, we can think of $X_2$ fibered over four copies of $X_1$ themselves fibered over two copies of $X_0$ as a dependent type. We have a priori a dependency over eight copies of $X_0$ but these are actually four due to the coherence conditions on the functor (e.g. that $\delta^R \circ \delta^{R\kstar} = \delta^{(R\kstar)\circ(R)} = \delta^{RR} = \delta^{(\kstar R)\circ(R)} = \delta^R \circ \delta^{\kstar R}$, etc.). Based on these coherence conditions, the types of $X_0, X_1, X_2$ happen to have the form:
\begin{align*}
  X_0 & :              & \U                                                     \\
  X_1 & :              & X_0 \rightarrow X_0 \rightarrow \U                     \\
  X_2 & : \Pi a b c d. & X_1(a)(b) \rightarrow X_1(a)(c) \rightarrow X_1 (b)(d) \\
      &                & \rightarrow X_1 (c)(d) \rightarrow \U                  \\
  \ldots
\end{align*}

@@

But first, we start with a section gradually and informally explaining
the intuition of the construction. It can be read in parallel with the
tables if wished.

@@

We assign every $X_n$ the type $\fullframe[n] \rightarrow \U$ uniformly, applying to the description above the isomorphism between $A \rightarrow B \rightarrow C$ and $A \times B \rightarrow C$, or between $\Pi a: A.\, (B a \rightarrow C)$ and $(\Sigma a: A.\, B a) \rightarrow C$. In particular, $\fullframe[n]$ is a ``telescope'', i.e. a nesting of $\Sigma$-type. Let us now illustrate how we build $\fullframe[n]$. To begin, let us set $\fullframe[0] = \unittype$, so that the type $\U$ of $X_0$ can be equivalently formulated as $\unittype \rightarrow \U$. Then, more generally, we let each $\fullframe[n]$ consist of $n$ layers, written $\layer[n][p]$ with $p < n$, that we stack in order, starting from the $\unittype$ type and writing $\framep[n][p]$ for the $p$ first layers of a $\fullframe[n]$, so that $\fullframe[n]$ is $\framep[n][n]$. For instance, $X_1$ is made of one layer, so that it can be written as a $\Sigma$-type of an inhabitant of the $\unittype$ and $\layer[1][0]$, as shown in Fig.~\ref{fig:intuition}. Fig.~\ref{fig:intuition} also mentions how the type of $X_2$ is structured.

@@

So far, we have not paid attention to the fact that we have a dependent type, shown as a $\Sigma$. Let us be more precise about this requirement. First, $\fullframe[n]$ depends on all $X_i$ up to $n - 1$. So, we need to package up $X_i$, for $i < n$, into a nesting of $\Sigma$-type, which we abbreviate as $\X[][<n][]$. This allows us to give the type $\X[][<n][] \rightarrow \U$ to $\fullframe[n]$. Then, for $D: \X[][<n][]$, representing an initial prefix of $X_0, X_1, \ldots X_{n - 1}$, the indexed set $X_n$ has type $\fullframe[n](D) \rightarrow \U$. Thus, $\framep[n][p]$, $\layer[n][p]$ and $\painting[n][p]$ also depend on $D$. We can then refine the previous equation by showing the dependencies on $D$. In particular, $X_n$ is just $D.\tl$.
\begin{align*}
   & \framep[n][p](D) \defeq                                                                                      \\
   & \;\;\;\;\Sigma a_n: (\ldots (\Sigma \unitpoint: \unittype.\, \layer[n][0](D)) \ldots) .\,\layer[n][p - 1](D) \\
   & \painting[n][p](D) \defeq                                                                                    \\
   & \;\;\;\;\Sigma l_p: \layer[n][p](D).\, (\ldots (\Sigma l_n: \layer[n][n - 1](D) .\, D.\tl) \ldots)
\end{align*}

@@

The definition is dispatched over tables \ref{tab:coind}, \ref{tab:core}, \ref{tab:frames}, \ref{tab:faces} and~\ref{tab:coh}. Table \ref{tab:coind} describes a $\nu$-set in indexed form, as a coinductive limit of $n$-truncated $\nu$-sets. The $n$-truncated $\nu$-sets are themselves described in \ref{tab:core}. This is a nested $\Sigma$-type of which the $n$th component is a family over a $\fullframe$. The type $\fullframe$ is recursively defined in \ref{tab:frames}, using the auxiliary definitions of $\layer$ and $\painting$. The type $\layer$ and $\painting$ are dependent over $\restrf$, and these restrictions are defined on \ref{tab:faces}. These restrictions are defined using auxiliary definitions of $\restrl$ and $\restrc$.

Notably, the definition of $\restrl$ relies on a definitional equality expressing the commutation of the composition of $\restrf$. This commutation is not provable by computation so we have to prove it propositionally before using the reflection rule. Proving this itself requires an induction on the dimension, and on the structure of $\framep$, $\layer$, and $\painting$. This is what $\cohf$ proves, as shown in the table \ref{tab:coh}, using auxiliary definitions $\cohl$ and $\cohc$. Even though it looks independent of the other tables, $\cohf$ has to be proved mutually with the definitions of $\framep$, $\layer$, $\painting$, and their corresponding restrictions. More precisely, for a fixed $n$, the block of $\framep$, $\restrf$, and $\cohf$ has to be mutually defined by induction on $p$. Also, each of $\painting$, $\restrc$, and $\cohc$ is built by induction from $p$ to $n$. The $\painting$ block at $n$ relies on the $\framep$ block at $n$, but, the converse dependency is only on lower $n$, so this is well-founded. Note that $\layer$, $\restrl$ and $\cohl$ are just abbreviations. We leave however implicit the exact way this mutual recursion can be formalized at this stage of the paper.
